---
title: "EDS 232: Machine Learning in Environmental Science"
bibliography: ["ml-env.bib"]
output:
  distill::distill_article:
    pandoc_args: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

> “Everything that civilisation has to offer is a product of human intelligence; we cannot predict what we might achieve when this intelligence is magnified by the tools that AI may provide, but the eradication of war, disease, and poverty would be high on anyone’s list. Success in creating AI would be the biggest event in human history. Unfortunately, it might also be the last.”
-- Stephen Hawking.

# Course Schedule

- **Lectures/Labs**: MW 9:30am - 10:45am on Zoom

```{r}
source(here::here("_functions.R"))
get_sched() %>% 
  dt_sched()
```

# Course Description

Machine learning can help process big/complex data and extract knowledge. It forms one of the foundations in data science. This course provides a broad introduction to machine learning and statistical pattern recognition. Topics include supervised learning (decision tree, random forest, support vector machines, neural networks) and unsupervised learning (clustering, dimensionality reduction, deep learning). Problems and exercises are framed within environmental science applications. The course will use programming languages like R and Python to support learning how to do advanced scientific programming to solve real environmental problems.

# Instructor

Ben Best (ben\@ecoquants.com)

- **Office hours**: Tuesdays TBD on Zoom

# Readings

```{r, eval=F}
library(citr)
md_cite("Bennett", bib_file="eds211-team.bib")
```

- Week 1:
  * @zhongMachineLearningNew2021
  * @elithSpeciesDistributionModels2009

