---
title: 'Lab 1d. Species Distribution Modeling - Evaluate Models'
editor_options: 
  chunk_output_type: console
---

# Learning Objectives {.unnumbered}

Now you'll complete the modeling workflow with the steps to **evaluate** model performance and **calibrate** model parameters.

```{r fig.cap="Full model workflow with calibrate and evaluate steps emphasized.", echo=F}
# [model-workflow_evaluate-calibrate - Google Drawings](https://docs.google.com/drawings/d/10hr0L_Iu_2DleLHzQmQWJGocG87yRbSX8jHne7fiZr0/edit)
knitr::include_graphics("https://docs.google.com/drawings/d/e/2PACX-1vSO_TOz2eZ_qmdIrsrXS1If82bUv1Sno1U5bJ2y8GD7S1PhrrSYMeKCxKy6GtSZp6NC01tBZfUnXPdD/pub?w=933&h=354")
```

# Setup

```{r setup}
# global knitr chunk options
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE)

# load packages
librarian::shelf(
  dismo, # species distribution modeling: maxent(), predict(), evaluate(), 
  dplyr, ggplot2, GGally, here, maptools, readr, 
  raster, readr, rsample, sf,
  usdm)  # uncertainty analysis for species distribution models: vifcor()
select = dplyr::select

# options
set.seed(42)
options(
  scipen = 999,
  readr.show_col_types = F)
ggplot2::theme_set(ggplot2::theme_light())

# paths
dir_data      <- here("data/sdm")
pts_geo       <- file.path(dir_data, "pts.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")
mdl_maxv_rds  <- file.path(dir_data, "mdl_maxent_vif.rds")

# read points of observation: presence (1) and absence (0)
pts <- read_sf(pts_geo)

# read raster stack of environment
env_stack <- raster::stack(env_stack_grd)
```

## Split observations into training and testing

```{r dt-data-prereq, echo=TRUE}
# create training set with 80% of full data
pts_split  <- rsample::initial_split(
  pts, prop = 0.8, strata = "present")
pts_train  <- rsample::training(pts_split)
pts_test   <- rsample::testing(pts_split)

pts_train_p <- pts_train %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_train_a <- pts_train %>% 
  filter(present == 0) %>% 
  as_Spatial()
```

# Calibrate: Model Selection

```{r}
# show pairs plot before multicollinearity reduction with vifcor()
pairs(env_stack)
```

```{r}
# calculate variance inflation factor per predictor, a metric of multicollinearity between variables
vif(env_stack)

# stepwise reduce predictors, based on a max correlation of 0.7 (max 1)
v <- vifcor(env_stack, th=0.7) 
v
# reduce enviromental raster stack by 
env_stack_v <- usdm::exclude(env_stack, v)

# show pairs plot after multicollinearity reduction with vifcor()
pairs(env_stack_v)
```

```{r}
# fit a maximum entropy model
if (!file.exists(mdl_maxv_rds)){
  mdl_maxv <- maxent(env_stack_v, sf::as_Spatial(pts_train))
  readr::write_rds(mdl_maxv, mdl_maxv_rds)
}
mdl_maxv <- read_rds(mdl_maxv_rds)

# plot variable contributions per predictor
plot(mdl_maxv)

# plot term plots
response(mdl_maxv)

# predict
y_maxv <- predict(env_stack, mdl_maxv) #, ext=ext, progress='')

plot(y_maxv, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

# Evaluate: Model Performance

## Area Under the Curve (AUC), Reciever Operater Characteristic (ROC) Curve and Confusion Matrix

```{r}
pts_test_p <- pts_test %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_test_a <- pts_test %>% 
  filter(present == 0) %>% 
  as_Spatial()

y_maxv <- predict(mdl_maxv, env_stack)
#plot(y_maxv)

e <- dismo::evaluate(
  p     = pts_test_p,
  a     = pts_test_a, 
  model = mdl_maxv,
  x     = env_stack)
e

plot(e, 'ROC')

thr <- threshold(e)[['spec_sens']]
thr
p_true <- na.omit(raster::extract(y_maxv, pts_test_p) >= thr)
a_true <- na.omit(raster::extract(y_maxv, pts_test_a) < thr)

# (t)rue/(f)alse (p)ositive/(n)egative rates
tpr <- sum(p_true)/length(p_true)
fnr <- sum(!p_true)/length(p_true)
fpr <- sum(!a_true)/length(a_true)
tnr <- sum(a_true)/length(a_true)

matrix(
  c(tpr, fnr,
    fpr, tnr), 
  nrow=2, dimnames = list(
    c("present_obs", "absent_obs"),
    c("present_pred", "absent_pred")))

# add point to ROC plot
points(fpr, tpr, pch=23, bg="blue")

plot(y_maxv > thr)
```

# Lab1 Submission

To submit Lab 1, please submit the path on `taylor.bren.ucsb.edu` to your single consolidated Rmarkdown document (i.e. combined from seperate parts) that you successfully knitted here:

-   [Submit Lab 1. Species Distribution Modeling](https://forms.gle/u9ZC7rvkcQMojxv3A)

The path should start `/Users/` and end in `.Rmd`. Please be sure to have succesfully knitted to html, as in I should see the `.html` file there by the same name.

In your lab, please be sure to include:

+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| \# | Lab          | Item                                                                                                                                                                                      | Points |
+====+==============+===========================================================================================================================================================================================+=======:+
| 1  | 1a.Explore   | Name \_Genus species\_ used to search GBIF for observations. (0.5 pts)                                                                                                                    | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 2  | 1a.Explore   | Image of your species. I like to see wildlife.                                                                                                                                            | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 3  | 1a.Explore   | Map of distribution of points                                                                                                                                                             | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 4  | 1a.Explore   | Question: How many observations total are in GBIF for your species?                                                                                                                       | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 5  | 1a.Explore   | Question: Did you have to perform any data cleaning steps?                                                                                                                                | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 6  | 1a.Explore   | Question: What environmental layers did you choose as predictors? Can you find any support for these in the literature?                                                                   | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 7  | 1a.Explore   | Raster plot of environmental rasters clipped to species range                                                                                                                             | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 8  | 1a.Explore   | Map with pseudo-absence points                                                                                                                                                            | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 9  | 1a.Explore   | Term plots                                                                                                                                                                                | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 10 | 1b.Regress   | Plot of ggpairs                                                                                                                                                                           | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 11 | 1b.Regress   | Linear model output with range of values                                                                                                                                                  | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 12 | 1b.Regress   | Generalized Linear Model (GLM) output                                                                                                                                                     | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 13 | 1b.Regress   | GLM term plots                                                                                                                                                                            | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 14 | 1b.Regress   | Generalized Additive Model (GAM) output                                                                                                                                                   | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 15 | 1b.Regress   | GAM term plots                                                                                                                                                                            | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 16 | 1b.Regress   | Question: Which GAM environmental variables, and even range of values, seem to contribute most towards presence (above 0 response) versus absence (below 0 response)?                     | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 17 | 1b.Regress   | Maxent model output                                                                                                                                                                       | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 18 | 1b.Regress   | Maxent variable contribution plot                                                                                                                                                         | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 19 | 1b.Regress   | Maxent term plots                                                                                                                                                                         | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 20 | 1b.Regress   | Question: Which Maxent environmental variables, and even range of values, seem to contribute most towards presence (closer to 1 response) and how might this differ from the GAM results? | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 21 | 1c.Trees     | Tabular counts of 1 vs 0 before and after split                                                                                                                                           | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 22 | 1c.Trees     | Rpart model output and plot, depth=1                                                                                                                                                      | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 23 | 1c.Trees     | Rpart model output and plot, depth=default                                                                                                                                                | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 24 | 1c.Trees     | Rpart complexity parameter plot                                                                                                                                                           | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 25 | 1c.Trees     | Question: Based on the complexity plot threshold, what size of tree is recommended?                                                                                                       | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 26 | 1c.Trees     | Rpart variable importance plot                                                                                                                                                            | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 27 | 1c.Trees     | Question: what are the top 3 most important variables of your model?                                                                                                                      | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 28 | 1c.Trees     | RandomForest variable importance                                                                                                                                                          | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 29 | 1c.Trees     | Question: How might variable importance differ between rpart and RandomForest in your model outputs?                                                                                      | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 30 | 1d.Evaluate  | Plot of pairs from environmental stack                                                                                                                                                    | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 31 | 1d.Evaluate  | VIF per variable                                                                                                                                                                          | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 32 | 1d.Evaluate  | Variables after VIF collinearity removal                                                                                                                                                  | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 33 | 1d.Evaluate  | Plof of pairs after VIF collinearity removal                                                                                                                                              | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 34 | 1d.Evaluate  | Plot of variable contribution                                                                                                                                                             | 0.3    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 35 | 1d.Evaluate  | Question: Which variables were removed due to multicollinearity and what is the rank of most to least important remaining variables in your model?                                        | 1      |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 36 | 1d.Evaluate  | Maxent term plots                                                                                                                                                                         | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 37 | 1d.Evaluate  | Map of Maxent prediction                                                                                                                                                                  | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 38 | 1d.Evaluate  | ROC threshold value maximizing specificity and sensitivity                                                                                                                                | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 39 | 1d.Evaluate  | Confusion matrix with percentages                                                                                                                                                         | 0.5    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 40 | 1d.Evaluate  | AUC plot                                                                                                                                                                                  | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+
| 41 | 1d.Evaluate  | Map of binary habitat                                                                                                                                                                     | 0.4    |
+----+--------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------+

