---
title: "Lab 2a. Community - Ordinate"
author: "Ben Best"
date: "1/24/2022"
output: html_document
bibliography: ["ml-env.bib"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives {.unnumbered}

In this lab, you will play with **unsupervised classification** techniques while working with **ecological community** datasets.

- **Ordination** orders sites near each other based on similarity. It is a multivariate analysis technique used to effectively collapse dependent axes into fewer dimensions, i.e. dimensionality reduction.

  - **Principal Components Analyses (PCA)** is the most common and oldest technique that assumes linear relationships between axes.
  
  - **Non-metric MultiDimensional Scaling (NMDS)** allows for non-linear relationships.
  
# Ordination

## Principal Components Analysis (PCA)

```{r}
# load R packages
librarian::shelf(
  dplyr, ggplot2, h2o)

# set seed for reproducible results
set.seed(42)

# get data
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"
my_basket <- readr::read_csv(url)
dim(my_basket)  

h2o.no_progress()  # turn off progress bars for brevity
h2o.init(max_mem_size = "5g")  # connect to H2O instance

# convert data to h2o object
my_basket.h2o <- as.h2o(my_basket)

# run PCA
my_pca <- h2o.prcomp(
  training_frame = my_basket.h2o,
  pca_method = "GramSVD",
  k = ncol(my_basket.h2o), 
  transform = "STANDARDIZE", 
  impute_missing = TRUE,
  max_runtime_secs = 1000)
my_pca

my_pca@model$eigenvectors %>% 
  as.data.frame() %>% 
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, reorder(feature, pc1))) +
  geom_point()

my_pca@model$eigenvectors %>% 
  as.data.frame() %>% 
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, pc2, label = feature)) +
  geom_text()


# Compute eigenvalues
eigen <- my_pca@model$importance["Standard deviation", ] %>%
  as.vector() %>%
  .^2
  
# Sum of all eigenvalues equals number of variables
sum(eigen)
## [1] 42

# Find PCs where the sum of eigenvalues is greater than or equal to 1
which(eigen >= 1)

# Extract PVE and CVE
ve <- data.frame(
  PC  = my_pca@model$importance %>% seq_along(),
  PVE = my_pca@model$importance %>% .[2,] %>% unlist(),
  CVE = my_pca@model$importance %>% .[3,] %>% unlist())

# Plot PVE and CVE
ve %>%
  tidyr::gather(metric, variance_explained, -PC) %>%
  ggplot(aes(PC, variance_explained)) +
  geom_point() +
  facet_wrap(~ metric, ncol = 1, scales = "free")

# How many PCs required to explain at least 75% of total variability
min(which(ve$CVE >= 0.75))

# Screee plot criterion
data.frame(
  PC  = my_pca@model$importance %>% seq_along,
  PVE = my_pca@model$importance %>% .[2,] %>% unlist()) %>%
  ggplot(aes(PC, PVE, group = 1, label = PC)) +
  geom_point() +
  geom_line() +
  geom_text(nudge_y = -.002)
```


## Non-metric MultiDimensional Scaling (NMDS)

### Unconstrained Ordination on Species

- vegantutor.pdf: 2.1 Non-metric Multidimensional scaling

```{r}
# load R packages
librarian::shelf(
  vegan, vegan3d)

# vegetation and environment in lichen pastures from Vare et al (1995)
data("varespec") # species
data("varechem") # chemistry

data(varespec)
vare.dis <- vegdist(varespec)
vare.mds0 <- monoMDS(vare.dis)
stressplot(vare.mds0)

ordiplot(vare.mds0, type = "t")

vare.mds <- metaMDS(varespec, trace = FALSE)
vare.mds

plot(vare.mds, type = "t")
```


### Overlay with Environment

- vegantutor.pdf: 
  * 3 Environmental interpretation
  * 3.1 Vector fitting
  * 3.2 Surface fitting

```{r}
ef <- envfit(vare.mds, varechem, permu = 999)
ef

plot(vare.mds, display = "sites")
plot(ef, p.max = 0.05)


ef <- envfit(vare.mds ~ Al + Ca, data = varechem)
plot(vare.mds, display = "sites")
plot(ef)

tmp <- with(varechem, ordisurf(vare.mds, Al, add = TRUE))
ordisurf(vare.mds ~ Ca, data=varechem, add = TRUE, col = "green4")
```


### Constrained Ordination on Species and Environment

- vegantutor.pdf: 
  * 4 Constrained ordination
  * 4.1 Model specification

```{r}
# ordinate on species constrained by three soil elements
vare.cca <- cca(varespec ~ Al + P + K, varechem)
vare.cca

# plot ordination
plot(vare.cca)

# plot 3 dimensions
ordiplot3d(vare.cca, type = "h")
  
if (interactive()){
  ordirgl(vare.cca)
}
```

