---
title: "Lab 2b. Community - Ordination"
author: "Ben Best"
date: "1/24/2022"
output: html_document
bibliography: ["ml-env.bib"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives {.unnumbered}

In this lab, you will play with **unsupervised classification** techniques while working with **ecological community** datasets.

- **Ordination** orders sites near each other based on similarity. It is a multivariate analysis technique used to effectively collapse dependent axes into fewer dimensions, i.e. dimensionality reduction.

  - **Principal Components Analyses (PCA)** is the most common and oldest technique that assumes linear relationships between axes. You will follow a non-ecological example from [Chapter 17 Principal Components Analysis | Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/pca.html) to learn about this commonly used technique.
  
  - **Non-metric MultiDimensional Scaling (NMDS)** allows for non-linear relationships. This ordination technique is implemented in the R package [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html). You'll use an ecological dataset, species and environment from lichen pastures that reindeer forage upon, with excerpts from the [vegantutor vignette](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf) ([source](https://github.com/jarioksa/vegandocs)) to apply these techniques:
    - **Unconstrained ordination** on species using NMDS;
    - Overlay with environmental gradients; and
    - **Constrained ordination** on species and environmnent using another ordination technique, **canonical correspondence analysis (CCA)**.
  
# Ordination

Ordination 

## Principal Components Analysis (PCA)

Although this example uses a non-ecological dataset, it goes through the materials walk through the idea and procedure of conducting an ordination using the most widespread technique.

Please read the entirety of [Chapter 17 Principal Components Analysis | Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/pca.html#finding-principal-components). Supporting text is mentioned below where code is run.

### Prerequisites

See supporting text: [17.1 Prerequisites](https://bradleyboehmke.github.io/HOML/pca.html#prerequisites-14)

```{r}
# load R packages
librarian::shelf(
  dplyr, ggplot2, h2o)

# set seed for reproducible results
set.seed(42)

# get data
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"
my_basket <- readr::read_csv(url)
dim(my_basket)

my_basket
```

From [Section 1.4](https://bradleyboehmke.github.io/HOML/intro.html#data):

- `my_basket.csv`: Grocery items and quantities purchased. Each observation represents a single basket of goods that were purchased together.
  * Problem type: unsupervised basket analysis
  * response variable: NA
  * features: 42
  * observations: 2,000
  * objective: use attributes of each basket to identify common groupings of items purchased together.

### Performing PCA in R

See supporting text: [17.4 Performing PCA in R](https://bradleyboehmke.github.io/HOML/pca.html#performing-pca-in-r)

```{r}
h2o.no_progress()  # turn off progress bars for brevity
h2o.init(max_mem_size = "5g")  # connect to H2O instance

# convert data to h2o object
my_basket.h2o <- as.h2o(my_basket)

# run PCA
my_pca <- h2o.prcomp(
  training_frame = my_basket.h2o,
  pca_method = "GramSVD",
  k = ncol(my_basket.h2o), 
  transform = "STANDARDIZE", 
  impute_missing = TRUE,
  max_runtime_secs = 1000)
my_pca

my_pca@model$eigenvectors %>% 
  as.data.frame() %>% 
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, reorder(feature, pc1))) +
  geom_point()

my_pca@model$eigenvectors %>% 
  as.data.frame() %>% 
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, pc2, label = feature)) +
  geom_text()
```

### Eigenvalue criterion

See supporting text: [17.5.1 Eigenvalue criterion](https://bradleyboehmke.github.io/HOML/pca.html#eigenvalue-criterion).


```{r}
# Compute eigenvalues
eigen <- my_pca@model$importance["Standard deviation", ] %>%
  as.vector() %>%
  .^2
  
# Sum of all eigenvalues equals number of variables
sum(eigen)
## [1] 42

# Find PCs where the sum of eigenvalues is greater than or equal to 1
which(eigen >= 1)

# Extract PVE and CVE
ve <- data.frame(
  PC  = my_pca@model$importance %>% seq_along(),
  PVE = my_pca@model$importance %>% .[2,] %>% unlist(),
  CVE = my_pca@model$importance %>% .[3,] %>% unlist())

# Plot PVE and CVE
ve %>%
  tidyr::gather(metric, variance_explained, -PC) %>%
  ggplot(aes(PC, variance_explained)) +
  geom_point() +
  facet_wrap(~ metric, ncol = 1, scales = "free")

# How many PCs required to explain at least 75% of total variability
min(which(ve$CVE >= 0.75))

# Screee plot criterion
data.frame(
  PC  = my_pca@model$importance %>% seq_along,
  PVE = my_pca@model$importance %>% .[2,] %>% unlist()) %>%
  ggplot(aes(PC, PVE, group = 1, label = PC)) +
  geom_point() +
  geom_line() +
  geom_text(nudge_y = -.002)
```


## Non-metric MultiDimensional Scaling (NMDS)

### Unconstrained Ordination on Species

See supporting text: **2.1 Non-metric Multidimensional scaling** in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf): 

```{r}
# load R packages
librarian::shelf(
  vegan, vegan3d)

# vegetation and environment in lichen pastures from Vare et al (1995)
data("varespec") # species
data("varechem") # chemistry

varespec %>% tibble()
vare.dis <- vegdist(varespec)
vare.mds0 <- monoMDS(vare.dis)
stressplot(vare.mds0)

ordiplot(vare.mds0, type = "t")

vare.mds <- metaMDS(varespec, trace = FALSE)
vare.mds

plot(vare.mds, type = "t")
```


### Overlay with Environment

See supporting text in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf): 
  * 3 Environmental interpretation
  * 3.1 Vector fitting
  * 3.2 Surface fitting

```{r}
ef <- envfit(vare.mds, varechem, permu = 999)
ef

plot(vare.mds, display = "sites")
plot(ef, p.max = 0.05)


ef <- envfit(vare.mds ~ Al + Ca, data = varechem)
plot(vare.mds, display = "sites")
plot(ef)

tmp <- with(varechem, ordisurf(vare.mds, Al, add = TRUE))
ordisurf(vare.mds ~ Ca, data=varechem, add = TRUE, col = "green4")
```


### Constrained Ordination on Species and Environment

See supporting text in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf): 
  * 4 Constrained ordination
  * 4.1 Model specification
  
Technically, this uses another technique `cca`, or canonical correspondence analysis.

```{r}
# ordinate on species constrained by three soil elements
vare.cca <- cca(varespec ~ Al + P + K, varechem)
vare.cca

# plot ordination
plot(vare.cca)

# plot 3 dimensions
ordiplot3d(vare.cca, type = "h")
  
if (interactive()){
  ordirgl(vare.cca)
}
```

